Bayesian Restricted Likelihood Methods: Conditioning on Insufficient Statistics in Bayesian Regression\thanksref{T1}

### 2020-05-09 revision_2/paperV2.tex
Steve's edits from May 7th

\response{The advantages and disadvantages of the method are detailed throughout the paper using simulated data and real data. One conceptual advantage is that inferences and predictions are less sensitive to features of the data not captured by
the conditioning statistics. Choosing statistics targeting main features of interest allows for more targeted inference on these features. The analysis can help to better understand other features, such as outliers, not capture by the conditioning statistics.} 
\response{The examples in the paper suggest advantages in situations where outliers are concern and there is significant prior information for the non-outlying portion of the model that is not outweighed by the data. The main disadvantage over traditional robust estimate techniques are mainly computational. In Section ~\ref{BayesLinMod} we detail a data-augmented MCMC algorithm to fit the models proposed in this paper. This requires an additional computational step for each iteration of the chain. Details of the additional burden are given and one must weigh the advantages of these methods with this additional burden. Since it is typically that the restricted likelihood posterior converges to the same expected value as the conditioning statistics as the sample size grows, it may not be practical to implement our method in all situations.}

@@

\subsection{The Bayesian linear model}
We focus on the use of restricted likelihood for the Bayesian linear
model with a standard formulation: 
\begin{eqnarray}
\label{LinearModel}
\bth&=&(\bbeta,\sigma^2) \sim  \pi(\bth) 
\nonumber\\
y_i  & =  & x_i^\top \bbeta + \epsilon_i , \mbox{ for } i = 1, \ldots, n 
\end{eqnarray}
where $x_i$ and $\bbeta \in \mathbb{R}^p$, $\sigma^2 \in \mathbb{R}^+$, 
and the $\epsilon_i$ are independent draws from a distribution with center $0$ and scale $\sigma$. $X$ denotes the design matrix whose rows are  $x_i^\top$. For the restricted likelihood model,  conditioning statistics are assumed to be of the form $T(\by) = (\bb(X, \by), s(X, \by))$ where $\bb(X, \by)= (b_1(X,\by), \dots,b_p(X,\by))^\top\in \mathbb R^{p}$ is an estimator for the regression coefficients and $s(X, \by)\in \{0\} \cup {\mathbb R}^+$ is an estimator of the scale. Throughout, observed data and summary statistic is denoted by $\by_{obs}$ and $T(\by_{obs})=(\bb(X, \by_{obs}), s(X, \by_{obs}))$, respectively. 
Several conditions are imposed on the model and statistic to ensure validity of the MCMC algorithm:
\begin{itemize}
\labitem{C1}{fullRank} The $n \times p$ design matrix, $X$, whose $i^{th}$ row is $x_i^\top$, 
is of full column rank.  
\labitem{C2}{supReal} The $\epsilon_i$ are a random sample from some distribution which has a density with 
respect to Lebesgue measure on the real line and for which the support is the real line.  
\labitem{C3}{asb}$\bb(X,\by)$ is almost surely continuous and differentiable with respect to $\by$.  
\labitem{C4}{as} $s(X,\by)$ is almost surely positive, continuous, and differentiable with respect to $\by$.  
\labitem{C5}{regEq} $\bb(X,\by+X\bv)=\bb(X,\by)+\bv \ \ \text{for  all}\ \bv\in\mathbb{R}^p$. 
\labitem{C6}{scaleEqReg} $\bb(X,a\by)=a\bb(X,\by)\ \ \ \text{for all constants } a$.  
\labitem{C7}{regIn} $s(X,\by+X\bv)=s(X,\by) \ \ \text{for all}\ \bv\in\mathbb{R}^p$.  
\labitem{C8}{scaleEq2Reg} $s(X, a\by)=|a|s(X,\by) \ \ \text{for all constants } a$.  
\end{itemize}
Properties \ref{regEq} and \ref{scaleEqReg} of $\bb$ are called
\textit{regression} and \textit{scale equivariance},
respectively.  Properties \ref{regIn} and \ref{scaleEq2Reg} of $s$ are called \textit{regression invariance}
and \textit{scale equivariance}. 
Many estimators satisfy the above properties, including several traditional simultaneous M-estimators \citep{huber2009, maronna2006} for which the \texttt{R} package \texttt{brlm} (\texttt{github.com/jrlewi/brlm}) is available to implement the MCMC described here. \response{These M-estimators satisfy \ref{asb}-\ref{as} since they are optimizers of (almost surely) continuous and differentiable objective functions. Constraints  \ref{regEq}-\ref{scaleEq2Reg} are often satisfied by location and scale estimators but should be checked on a case by case basis.}  More software development is required to extend the MCMC implementation beyond the M-estimators discussed here. The current version of the \texttt{R} package also implements the direct computational methods described in \cite{lewis2014}. These methods are effective in lower dimensional problems and were used in both examples in Section \ref{illustrations}.

@@

\vskip 0.05 in
\noindent
{\bf The proposal density} \\
Putting all the pieces of the Jacobian together we have the following result. Any dependence on other variables, including current states in the Markov chain, is made implicit. 
\begin{theorem} 
Assume that conditions \ref{fullRank}-\ref{scaleEq2Reg} hold.  Let $\bz^{*}$ be sampled on the unit sphere in $\mc {C}^\perp (X)$ with density $p(\bz^{*})$.  Using the transformation of $\bz^*$ to $\by\in \mc A$ described in Theorem \ref{Transformation}, the density of $\by$ is
\begin{equation}
\label{dens:ystst}
p(\by)=p(\bz^*) r^{-(n-p-1)} \cos(\gamma)\text{Vol} (P)
\end{equation}
where $r={s(X,\boldsymbol{y}_{obs})}/{s(X,  \boldsymbol{z}^{*})}$,
and $\cos(\gamma)$ and $\text{Vol} (P)$ are as in equations \eqref{cosine} and \eqref{eq:volume}, respectively. 
\end{theorem} 
\response{The proposal is governed by the choice of $p(z^{*})$ and a poor choice could cause concern about the efficiency of the convergence of the MCMC algorithm. For all the examples in the paper we defined $p(z^{*})$ to simply be the uniform distribution on $\mathbb{S}$. The advantage of this choice is that it requires no further tuning parameters and we have noticed good mixing in terms of the ability of the chain to generate new data $\mb y$ that is accepted with reasonable probabilities. To implement in practice, we simply generate an n-dimensional independent standard normal $\mb y^{*}$ for the proposal and transform this via $h(\cdot)$. Theoretically, the random normal vector would be projected onto $\mc{C}^{\perp}(X)$ and scaled to unit norm to generate the uniform on $\mathbb{S}$. Using simple algebra and conditions \ref{regEq}-\ref{scaleEq2Reg}, one can show $h(\cdot)$ is invariant to this projection and scaling. Another option for the proposal suggested by a reviewer that the authors have yet to study is generating a random walk. As we are proposing values on a complex manifold, it might be possible to implement this by conducting the random walk on $\mb y^{*}$ before transforming via $h(\cdot)$. This could provide some advantages in some situations, though we have yet to run into any serious issues with convergence using the independent proposal we utilize here.}

@@

\response{Finally, it is clear the estimators themselves must be computed for every iteration of the Markov Chain. We have found this burden to be marginal in relation with respect to computing the needed Jacobian. In the simulations and real data analyses presented below, we will see that the additional burden is often of added value over traditional robust regression when substantial prior information is available that is not swamped by current data.}

@@

This paper develops a Bayesian version of restricted likelihood where posterior inference is conducted by conditioning on a summary statistic rather than the complete data.  The framework blends classical estimation with Bayesian methods.  
%Many routine choices in an analysis react to the gap between reality and the
%statistical model, where a bit of set-up work improves inferential
%performance.  Often, these choices can be recast in the framework of
%restricted likelihood presented here, lending them more formality and facilitating
%development of theoretical results. 
%A greater benefit of our framework is that it leads us to blend classical estimation with Bayesian methods. 
Here, we concentrate on outlier-prone settings where natural choices for the conditioning statistic are classical robust estimators targeting the mean of the non-outlying data (e.g., M-estimators).  The likelihood conditioned on these estimators is used to move from prior to posterior. The update follows Bayes' Theorem, conditioning on the observed estimators exactly.   Computation is driven by MCMC methods, requiring only a supplement to existing algorithms by adding a Gibbs step to sample from the space of data sets satisfying the observed statistic. This step has additional computation costs arising from the need to compute the estimator and an orthonormal basis derived from gradients of the estimator at each iteration. The cost of finding the basis can be reduced by exploiting properties of the geometric space from which the samples are drawn as described in Section~\ref{highDim}. We have seen good mixing of the MCMC chains across a wide-variety of examples. \response{We have found the additional computational burden to be beneficial in situation where  substantial prior information that will impact the results is available.}

>>>>

%\response{The advantages and disadvantages of the method are detailed throughout the paper using simulated data and real data. One conceptual advantage is that inferences and predictions are less sensitive to features of the data not captured by the conditioning statistics. Choosing statistics targeting main features of interest allows for more targeted inference on these features. The analysis can help to better understand other features, such as outliers, not capture by the conditioning statistics.} 
%\response{The examples in the paper suggest advantages in situations where outliers are concern and there is significant prior information for the non-outlying portion of the model that is not outweighed by the data. The main disadvantage over traditional robust estimate techniques are mainly computational. In Section ~\ref{BayesLinMod} we detail a data-augmented MCMC algorithm to fit the models proposed in this paper. This requires an additional computational step for each iteration of the chain. Details of the additional burden are given and one must weigh the advantages of these methods with this additional burden. Since it is typically that the restricted likelihood posterior converges to the same expected value as the conditioning statistics as the sample size grows, it may not be practical to implement our method in all situations.}
\response{The advantages and disadvantages of the method are detailed throughout the paper using simulated and real data. One conceptual advantage of our method is that inferences and predictions are less sensitive to features of the data not captured by
the conditioning statistics than are methods based on the complete likelihood. Choosing statistics targeting the main features of interest allows for inference that focuses on these features. The analysis can help to better understand other features which may not be captured by the conditioning statistics, such as outliers.}

\response{The examples in the paper provide a Bayesian analog of classical robust estimators.  The main disadvantage of our methods relative to the classical estimators is computational.  In Section~\ref{BayesLinMod} we detail a data-augmentation MCMC algorithm to fit the models proposed in this paper.  The advantages are those of Bayesian methods.  As is standard for Bayes-classical comparisons, the Bayesian method requires greater computational effort while providing better inference.  As a referee notes, asymptotically, the Bayesian and classical parameter estimates are often very close and have the same limiting posterior variance / sampling variance.  In situations where asymptotic approximation suffices, there is no need to use the computational techniques developed in this paper.}

@@

\subsection{The Bayesian linear model}
We focus on the use of restricted likelihood for the Bayesian linear
model with a standard formulation: 
\begin{eqnarray}
\label{LinearModel}
\bth&=&(\bbeta,\sigma^2) \sim  \pi(\bth) 
\nonumber\\
y_i  & =  & x_i^\top \bbeta + \epsilon_i , \mbox{ for } i = 1, \ldots, n 
\end{eqnarray}
where $x_i$ and $\bbeta \in \mathbb{R}^p$, $\sigma^2 \in \mathbb{R}^+$, 
and the $\epsilon_i$ are independent draws from a distribution with center $0$ and scale $\sigma$. $X$ denotes the design matrix whose rows are  $x_i^\top$. For the restricted likelihood model,  conditioning statistics are assumed to be of the form $T(\by) = (\bb(X, \by), s(X, \by))$ where $\bb(X, \by)= (b_1(X,\by), \dots,b_p(X,\by))^\top\in \mathbb R^{p}$ is an estimator for the regression coefficients and $s(X, \by)\in \{0\} \cup {\mathbb R}^+$ is an estimator of the scale. Throughout, observed data and summary statistic is denoted by $\by_{obs}$ and $T(\by_{obs})=(\bb(X, \by_{obs}), s(X, \by_{obs}))$, respectively. 
Several conditions are imposed on the model and statistic to ensure validity of the MCMC algorithm:
\begin{itemize}
\labitem{C1}{fullRank} The $n \times p$ design matrix, $X$, whose $i^{th}$ row is $x_i^\top$, 
is of full column rank.  
\labitem{C2}{supReal} The $\epsilon_i$ are a random sample from some distribution which has a density with 
respect to Lebesgue measure on the real line and for which the support is the real line.  
\labitem{C3}{asb}$\bb(X,\by)$ is almost surely continuous and differentiable with respect to $\by$.  
\labitem{C4}{as} $s(X,\by)$ is almost surely positive, continuous, and differentiable with respect to $\by$.  
\labitem{C5}{regEq} $\bb(X,\by+X\bv)=\bb(X,\by)+\bv \ \ \text{for  all}\ \bv\in\mathbb{R}^p$. 
\labitem{C6}{scaleEqReg} $\bb(X,a\by)=a\bb(X,\by)\ \ \ \text{for all constants } a$.  
\labitem{C7}{regIn} $s(X,\by+X\bv)=s(X,\by) \ \ \text{for all}\ \bv\in\mathbb{R}^p$.  
\labitem{C8}{scaleEq2Reg} $s(X, a\by)=|a|s(X,\by) \ \ \text{for all constants } a$.  
\end{itemize}
Properties \ref{regEq} and \ref{scaleEqReg} of $\bb$ are called
\textit{regression} and \textit{scale equivariance},
respectively.  Properties \ref{regIn} and \ref{scaleEq2Reg} of $s$ are called \textit{regression invariance}
and \textit{scale equivariance}. 
Many estimators satisfy the above properties, including several traditional simultaneous M-estimators \citep{huber2009, maronna2006} for which the \texttt{R} package \texttt{brlm} (\texttt{github.com/jrlewi/brlm}) is available to implement the MCMC described here. 
%\response{These M-estimators satisfy \ref{asb}-\ref{as} since they are optimizers of (almost surely) continuous and differentiable objective functions. Constraints  \ref{regEq}-\ref{scaleEq2Reg} are often satisfied by location and scale estimators but should be checked on a case by case basis.}
\response{These M-estimators satisfy \ref{abs} and \ref{as} since they are optimizers of continuous and differentiable objective functions. Constraints \ref{regEq}-\ref{scaleEq2Reg} are often satisfied by location and scale estimators but should be checked on a case by case basis.}  
More software development is required to extend the MCMC implementation beyond the M-estimators discussed here. The current version of the \texttt{R} package also implements the direct methods described in \cite{lewis2014}. These methods are effective in lower dimensional problems and were used in both examples in Section \ref{illustrations}.

@@

\vskip 0.05 in
\noindent
{\bf The proposal density} \\
Putting all the pieces of the Jacobian together we have the following result. Any dependence on other variables, including current states in the Markov chain, is made implicit. 
\begin{theorem} 
Assume that conditions \ref{fullRank}-\ref{scaleEq2Reg} hold.  Let $\bz^{*}$ be sampled on the unit sphere in $\mc {C}^\perp (X)$ with density $p(\bz^{*})$.  Using the transformation of $\bz^*$ to $\by\in \mc A$ described in Theorem \ref{Transformation}, the density of $\by$ is
\begin{equation}
\label{dens:ystst}
p(\by)=p(\bz^*) r^{-(n-p-1)} \cos(\gamma)\text{Vol} (P)
\end{equation}
where $r={s(X,\boldsymbol{y}_{obs})}/{s(X,  \boldsymbol{z}^{*})}$,
and $\cos(\gamma)$ and $\text{Vol} (P)$ are as in equations \eqref{cosine} and \eqref{eq:volume}, respectively. 
\end{theorem} 
%\response{The proposal is governed by the choice of $p(z^{*})$ and a poor choice could cause concern about the efficiency of the convergence of the MCMC algorithm. For all the examples in the paper we defined $p(z^{*})$ to simply be the uniform distribution on $\mathbb{S}$. The advantage of this choice is that it requires no further tuning parameters and we have noticed good mixing in terms of the ability of the chain to generate new data $\mb y$ that is accepted with reasonable probabilities. To implement in practice, we simply generate an n-dimensional independent standard normal $\mb y^{*}$ for the proposal and transform this via $h(\cdot)$. Theoretically, the random normal vector would be projected onto $\mc{C}^{\perp}(X)$ and scaled to unit norm to generate the uniform on $\mathbb{S}$. Using simple algebra and conditions \ref{regEq}-\ref{scaleEq2Reg}, one can show $h(\cdot)$ is invariant to this projection and scaling. Another option for the proposal suggested by a reviewer that the authors have yet to study is generating a random walk. As we are proposing values on a complex manifold, it might be possible to implement this by conducting the random walk on $\mb y^{*}$ before transforming via $h(\cdot)$. This could provide some advantages in some situations, though we have yet to run into any serious issues with convergence using the independent proposal we utilize here.}
\response{The proposal is governed by the choice of $p(z^{*})$ and a poor choice could lead to an inefficient MCMC algorithm. For all examples in this paper we defined $p(z^{*})$ to be the uniform distribution on $\mathbb{S}$. The advantage of this choice is that it requires no further tuning parameters.  We have noticed good mixing in terms of the ability of the chain to generate new data $\mb y$ that is accepted with a reasonable probability. To implement the method in practice, we generate an n-dimensional independent standard normal $\mb y^{*}$ for the proposal and transform this via $h(\cdot)$. Theoretically, the random normal vector would be projected onto $\mc{C}^{\perp}(X)$ and scaled to unit norm to generate the uniform on $\mathbb{S}$. Using simple algebra and conditions \ref{regEq}-\ref{scaleEq2Reg}, one can show that $h(\cdot)$ is invariant to this projection and scaling. Another option for the proposal suggested by a reviewer that the authors have yet to study is generating a random walk. As we are proposing values on a complicated manifold, it might be possible to implement this by conducting the random walk on $\mb y^{*}$ before transforming via $h(\cdot)$. This could provide advantages in some situations, though we have yet to run into any serious issues with convergence using the independence proposal we utilize here.}

@@

%\response{Finally, it is clear the estimators themselves must be computed for every iteration of the Markov Chain. We have found this burden to be marginal in relation with respect to computing the needed Jacobian. In the simulations and real data analyses presented below, we will see that the additional burden is often of added value over traditional robust regression when substantial prior information is available that is not swamped by current data.}
\response{Finally, it is clear the estimators themselves must be computed for every iteration of the Markov Chain. We have found this burden to be marginal relative to computation of the needed Jacobian. In the simulations and real data analyses presented below, we will see that the additional computational expense needed to fit the Bayesian model is often worthwhile, leading to better performance compared to traditional, non-Bayesian robust regression estimators.  This is most evident when substantive prior information is available and information in the data is limited.}

@@

This paper develops a Bayesian version of restricted likelihood where posterior inference is conducted by conditioning on a summary statistic rather than the complete data.  The framework blends classical estimation with Bayesian methods.  
%Many routine choices in an analysis react to the gap between reality and the
%statistical model, where a bit of set-up work improves inferential
%performance.  Often, these choices can be recast in the framework of
%restricted likelihood presented here, lending them more formality and facilitating
%development of theoretical results. 
%A greater benefit of our framework is that it leads us to blend classical estimation with Bayesian methods. 
Here, we concentrate on outlier-prone settings where natural choices for the conditioning statistic are classical robust estimators targeting the mean of the non-outlying data (e.g., M-estimators).  The likelihood conditioned on these estimators is used to move from prior to posterior. The update follows Bayes' Theorem, conditioning on the observed estimators exactly.   Computation is driven by MCMC methods, requiring only a supplement to existing algorithms by adding a Gibbs step to sample from the space of data sets satisfying the observed statistic. This step has additional computation costs arising from the need to compute the estimator and an orthonormal basis derived from gradients of the estimator at each iteration. The cost of finding the basis can be reduced by exploiting properties of the geometric space from which the samples are drawn as described in Section~\ref{highDim}. We have seen good mixing of the MCMC chains across a wide-variety of examples. %\response{We have found the additional computational burden to be beneficial in situation where  substantial prior information that will impact the results is available.}
\response{We have found the benefits of using our Bayesian technique to outweigh the additional computational burden (relative to a classical estimator) in the situation where substantive prior information that will impact the results is available.}
